Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
err

priors are good

@@

We made four important developments of the open population
N-mixture model proposed by Dail and Madsen that make the
model more applicable to long-term data such as is collected
by the North American Breeding Bird Survey. First, we have
reconciled the objectives of traditional state-space models
with open population N-mixture models by illustrating how
classical models of population dynamics can be embedded within
the framework. Second, we have demonstrated methods of
accounting for zero-inflation in the time-series. Third, we
have illustrated how additional random effects such as
observer-specific detection probabilities can be
accommodated. Fourth, we presented a Bayesian analysis of the
model, which makes it much easier to estimate parameters and
to fine tune the model for specific purposes.

@@

[paragraph on random effects]
[paragraph on Bayesian]
In spite of the new extensions we have proposed, several aspects of
the model could be improved. First, the precision with which the
parameters of the state process can be estimated ultimately depends
upon how well detection probability is estimated. When there is only a
single survey per primary period, the information about detection
probability comes from deviations from the parametric assumptions
about population dynamics. Thus, without direct information about
detection probability, the estimates will be determined by model based
assumptions. Furthermore, there are multiple components of detection
probability that should be accounted for to minimize bias and yield
valid estimates of population size. [Nichols EURING paper] .

@@

[paragraph on spatially-explicit dispersal]
[Other possible extensions we could mention:
Spatially hierarchical model (so can use individual stops)
Other models of variability in growth or recruitment (such as negative binomial)
Add density dependence to survival and recruitment models]
[Discuss causes of low p values (and other BBS results) here or in BBS
results?]

>>>>

Bayesian inference has several appealing features. First, it allows
direct probability statements to be made about a hypothesis given
data. Second, priors are cool. We can combine data from multiple
sources or use existing estimates of parameters as prior
distributions. MCMC has a bad rap for being slow, but actually, in the
case of state-space models it may be one of the few options for
estimation, and in fact may be faster than maximizing likelihoods.

Here is some BUGS code:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how open population $N$-mixture models overcome these
limitations. We extended this class of models in several important
ways, while keeping in mind the practical issues ...

We made four important developments of the open population
N-mixture model proposed by \citep{dail_madsen:2011} that make the
model more applicable to long-term data such as is collected
by the North American Breeding Bird Survey. First, we have
reconciled the objectives of traditional state-space models
with open population $N$-mixture models by illustrating how
classical models of population dynamics can be embedded within
the framework. Second, we have demonstrated methods of
accounting for zero-inflation in the time-series. Third, we
have illustrated how additional random effects such as
observer-specific detection probabilities can be
accommodated. Fourth, we presented a Bayesian analysis of the
model, which, makes it possible to incorporate prior information when
available, and in some cases, facilitates parameter estimation.

@@

[paragraph on random effects]

[paragraph on Bayesian]
The ability to fit this class of models using freely available MCMC
software offers non-statisticians a powerful means .

In spite of the new extensions we have proposed, several aspects of
the model could be improved. First, the precision with which the
parameters of the state process can be estimated ultimately depends
upon how well detection probability is estimated. When there is only a
single survey per primary period, the information about detection
probability comes from deviations from the parametric assumptions
about population dynamics. Thus, without direct information about
detection probability, the estimates will be determined by model based
assumptions. Furthermore, there are multiple components of detection
probability that should be accounted for to minimize bias and yield
valid estimates of population size \citep{nichols_etal:2009}.

@@

[paragraph on spatially-explicit dispersal]
[Other possible extensions we could mention:
Spatially hierarchical model (so can use individual stops)
Other models of variability in growth or recruitment (such as negative binomial)
Add density dependence to survival and recruitment models]
[Discuss causes of low p values (and other BBS results) here or in BBS
results?]
Our emphasis was on increasing the practical utility of this class of
models, and so we avoided several conceptually-interesting extensions
that we belive would be computationally prohibitive in many
cases. Nonetheless, we will discuss one---spatially-explict models of
immigration.
